Sequence Specific Modeling of \emph{E.~coli} Cell-Free Protein Synthesis

### 2017-04-06 Paper_v1.tex
Need a break ...

\subsection{Analysis of CFPS performance}
To better understand the performance of CFPS reactions, we predicted the productivity, energy efficiency and carbon yield for the cell free production of eight model proteins with and without amino acid supplementation (Fig. \ref{fig:Prof}).
The expression of each of these proteins was under a P70a promoter, with the exception of CAT which we assumed was expressed using a T7 promoter.
In all cases, CFPS was supplied with glucose, however, we also considered different scenarios for amino acid supplementation.
First, the CFPS reaction was supplied with amino acids, and the system was also allowed to synthesize amino acids from glucose (AA uptake and synthesis).
Next, the CFPS reaction was supplied with both glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AA uptake w/o synthesis).
This scenario was consistent with common cell free extract preparation protocols which often involve amino acid supplementation; thus, we would expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction.
Lastly, the system was not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (AA synthesis w/o uptake).
Eight proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10),
chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP),
prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental metabolic measurements (see supplemental materials).

@@

\subsubsection{Productivity}
The theoretical maximum mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prof}A).
%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.
This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
However, each individual protein had a different level of productivity.
For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity given the same resources.
A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information).
CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
The relative high productivity of CAT was due to its T7 promoter.
CAT on a P70a promoter frollowed the same trendline as the other proteins and had a productivity of about 8.8 $\mu$M/h.
Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a P70a promoter in CFPS, given the size of the protein.
%The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
% further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.

@@

\subsection{Sensitivity analysis}
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for CAT, a representative protein (Fig.~\ref{fig:SI}).
The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield.
In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation being the most significant.
As expected, this showed that the translation rate was instrumental for productivity, and should be the first step investigated during optimization, prior to examining transcription parameters.
Underwood and coworkers have also shown that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly luciferase by 5-fold in CFPS systems by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the cases with amino acid supplementation, but its significance increased when amino acids were not available.
This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
On the other hand, amino acid uptake only showed significance for the case where amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray). Error bars represent a 95\% CI on the sensitivity index.}
\label{fig:SI}
\end{figure}

@@

\subsection{Flux distribution}
To investigate the differences between the three cases, we compared the flux distributions for CAT production predicted by ssFBA simulations (Fig.~\ref{fig:flux}).
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./figs/Fig-5-FluxDistribition-Optimal.pdf}
\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation, for three different cases: (A) amino acid uptake and synthesis, (B) amino acid uptake without synthesis, and (C) amino acid synthesis without uptake. Mean flux across the ensemble, normalized to glucose uptake flux. Thick arrows indicate flux to or from amino acids.}
\label{fig:flux}
\end{figure}
In the cases supplied with amino acids (Fig.~\ref{fig:flux}A-B), glucose traveled to acetyl-coenzyme A that generated NADH for oxidative phosphorylation to energize CFPS.
Interestingly, in the first case with amino acid synthesis reactions, ssFBA relied on a combination of glucose and amino acids to power the system
Amino acids rather than glucose powered the TCA cycle via glutamate to alpha-ketoglutarate which traveled to oxaloacetic acid and pyruvate for additional amino acid biosynthesis.
In the case where amino acid synthesis reactions were blocked (Fig.~\ref{fig:flux}B), ubiquinone was generated via \textit{nuo} to power oxidative phosphorylation, instead of relying on the TCA cycle.
Once the energy requirements for transcription and translation were met, amino acids were taken from the media to assemble the protein of interest.
These first two cases where amino acids were available had similar performance with a correlation of 0.99 between their flux distributions and were comparable in terms of productivity, energy efficiency and carbon yield for all proteins.
%This is most likely due to the efficient utilization of carbon to energize CFPS without the burden of amino acid biosynthesis.
In the case where amino acids must be synthesized, there was a drop in the performance metrics of energy efficiency and carbon yield compared to the cases where amino acids are avialble.
The flux distribution had a correlation of 0.90 when compared to both cases where amino acids were avaialbe.
This drop in performance is due to the burden of synthesizing amino acids, which require NADPH.
This leads to the relatively high flux in the first step of the pentose phosphate pathway to generate NADPH, thus, less NADH is available for oxidative phosphorylation.
The performance metrics and sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
%Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.

@@

Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm})
For example, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.
A knockout of the Entner-Doudoroff pathway has no effect on the norm productivity of CAT compared to no knockouts (Fig.~\ref{fig:norm}A).
In addition, pairwise knockouts of Entner-Doudoroff and other subgroups (i.e. pentose phosphate pathway, cofactors, folate metabolism, etc.) in the network result in the same optimal solution of CAT productivity.
However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B); ssFBA will reroute the flux to optimize the objective function.
Interestingly, there are crucial subgroups that are detrimental to CAT productivity.
For example, a single group knockout of glycolysis/gluconoeogenesis or oxidative phosphorylation leads to no CAT production.
There were also cases of group knockouts that had no effect on productivity or its flux distribution such as isoleucine, leucine, and valine biosynthesis knocked out with histidine biosynthesis.
Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
In our case, ssFBA reached the same optimal solution of CAT productivity for 39.9\% of the pairwise group knockouts, but had a different flux distribution than compared to no knockouts.
To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.
It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
\begin{figure}[t!]
\includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribition-Analysis.pdf}
\caption{The norm effect of pairwise knockouts of subgroups in the cell-free network (A) CAT productivity norm compared to no knockouts in the experimentally constrained case. (B) Flux distribution norm compared to no knockouts in the experimentally constrained case.}
\label{fig:norm}
\end{figure}

Taken together, we developed a sequence specific constraints based modeling approach to evaluate the performance of synthetic circuits in an \emph{E.~coli} CFPS system for a range of different proteins and three different cases.
We have shown first principle predictions for protein production of deGFP and CAT in agreement with experimental measurements, under two different promoters and two different cell-free extract systems, with few adjustable parameters in the promoter models taken from literature.
This modeling approach suggested trends for productivity, energy efficiency and carbon yield as a function of carbon number.
Furthermore, global sensitivity analysis identified oxygen uptake as being instrumental for maintaining a high energy efficiency and carbon yield.
The translation rate was identified as the rate limiting step for productivity.
The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize energy efficiency and carbon yield.
In conclusion, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.

@@

\subsection*{Formulation and solution of the model equations.}
The sequence-specific flux balance analysis problem was formulated as a linear program:
\begin{equation}
 \begin{multlined}
	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{X} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
	\mathrm{Subject \; to:}
	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
\mathcal{L}_{i} \leq w_i \leq \mathcal{U}_{i}  \qquad i=1,2,\hdots,\mathcal{R}
 \end{multlined}
\end{equation}
where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective cost vector
and $\mathcal{L}_{i}$ and $\mathcal{U}_{i}$ denote the lower and upper bounds on flux $w_{i}$, respectively.
The transcription (T) and translation (X) stoichiometry was modeled based upon the template reactions of Allen and Palsson \cite{Allen:2003aa}:

\begin{table}[!]
\centering
    \caption{Transcription and translation template reactions for protein production.}
    \renewcommand{\arraystretch}{1.2}
    \resizebox{\textwidth}{!}{\begin{tabular}{llcl} \toprule
        \textbf{Description} & \multicolumn{3}{l}{\textbf{Template reaction}} \\ \toprule
        Transcription initiation & $G_\mathcal{P}+R_{T}$ & $\longrightarrow$& $G_{\mathcal{P}}^{*}$ \\
        Transcription ($w_{T}$) & $G_{\mathcal{P}}^{*}+\quad\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right)$ & $\longrightarrow$& $mRNA+G_{\mathcal{P}}+R_{T}+\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}} \eta_{k} \cdot PPi$ \\
        mRNA degradation &  $mRNA$ &$\longrightarrow$& \quad $\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left\{k\right\}MP$ \\
	Translation initiation & $mRNA+R_{X}$ &$\longrightarrow$& $R_{X}^{*}$ \\
        tRNA charging & $\alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big)$ &$\longrightarrow$& $\alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)$ \\
 	&&& $\qquad \qquad \qquad \qquad \quad  {j=1,2,\hdots,20}$\\
        Translation ($w_{X}$) &  $R_{X}^{*}+\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big)$ &$\longrightarrow$& $\mathcal{P}+R_{X}+mRNA$\\
&&& $\quad +\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)$ \\ \bottomrule
    \end{tabular}}
\label{tbl:parameters}
\end{table}
where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
$R_{T}$ denotes the concentration of RNA polymerase,
$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase (open complex),
$\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
$Pi$ denotes inorganic phosphate,
$R_{X}$ denotes the ribosome concentration,
$R_{X}^{*}$ denotes bound ribosome,
and $AA_{j}$ denotes $j^{th}$ amino acid.

% \begin{eqnarray*}
% {G}_{\mathcal{P}}+{R}_{T} &\longrightarrow& G_{\mathcal{P}}^{*} \\
% G_{\mathcal{P}}^{*}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right) &\overset{\rm TX}{\longrightarrow}& mRNA+G_{\mathcal{P}}+R_{T}+ \sum_{k\in\left\{A,C,G,U\right\}} \eta_{k} \cdot PPi\\
% mRNA &\longrightarrow& \sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP \\
% mRNA+R_{X} &\longrightarrow& R_{X}^{*} \\
% %\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP +\alpha_{j}\cdot H_{2}O &\longrightarrow& \alpha_{j}\cdot AA_{j}\text{-}tRNA_{j}+\\
% %&& \qquad \alpha_{j}\cdot AMP+\alpha_{j}\cdot PPi \qquad{j=1,2,\hdots,20}\\
% \alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big) &\longrightarrow& \alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)\\
% && \qquad \qquad \qquad \qquad {j=1,2,\hdots,20}\\
% R_{X}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big) &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+R_{X}+mRNA\\
% && \quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)
% \end{eqnarray*}

>>>>

\subsection{Analysis of CFPS performance}
To better understand the performance of CFPS reactions, we predicted the productivity, energy efficiency and carbon yield for the cell free production of eight model proteins with and without amino acid supplementation (Fig. \ref{fig:Prof}).
The expression of each of these proteins was under a P70a promoter, with the exception of CAT which we assumed was expressed using a T7 promoter.
In all cases, CFPS was supplied with glucose; however, we considered different scenarios for amino acid supplementation.
First, the CFPS reaction was supplied with amino acids, and simultaneously the system was also allowed to synthesize amino acids from glucose (AA uptake and synthesis).
Next, the CFPS reaction was supplied with both glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AA uptake w/o synthesis).
This scenario was consistent with common cell free extract preparation protocols which often involve amino acid supplementation; thus, we expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction in these cases.
Lastly, the cell free reaction was not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (AA synthesis w/o uptake).
Eight proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10),
chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP),
prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental metabolic measurements (see supplemental materials).

@@

\subsubsection{Productivity}
The theoretical maximum mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B). The theoretical maximum productivity was within a standard deviation with and without
amino acid supplementation for each protein, but varied significantly between proteins.
This suggested the cell free reaction had sufficient metabolic precursors to power CFPS and synthesize each protein of interest with similar productivities.
However, the theoretical productivity was inversely proportional to protein size;
for instance, BMP10 (XX aa) had a productivity of about 2.5 $\mu$M/h whereas CAT (ZZ aa) had a productivity of about 12 $\mu$M/h.
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
The productivity and protein size were related by the power-law relationship $\alpha\times(L)^{\beta}$, where $\alpha = 6.0\times 10^{6}$ and $\beta=-1.9$.
Interestingly, CAT did not obey the trendline; the relatively high productivity of CAT was due to its T7 promoter.
CAT expressed under a P70a promoter followed the power-law correlation with a productivity of approximately 8.8 $\mu$M/h.
Taken together, the maximum productivity of a cell free reaction was found to be inversely proportional to protein size,
following a power-law relationship for proteins expressed under a P70a promoter.

% Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a  in CFPS, given the size of the protein.
%The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
% further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.

@@

\subsection{Sensitivity analysis}
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI}).
The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield.
In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation being the most significant.
As expected, this showed that the translation rate was instrumental for productivity, and should be the first step investigated during optimization, prior to examining transcription parameters.
Underwood and coworkers have also shown that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly luciferase by 5-fold in CFPS systems by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the cases with amino acid supplementation, but its significance increased when amino acids were not available.
This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
On the other hand, amino acid uptake only showed significance for the case where amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray). Error bars represent a 95\% CI on the sensitivity index.}
\label{fig:SI}
\end{figure}

@@

\subsection{Optimal Metabolic Flux Distribution}
Amino acid supplementation altered the optimal metabolic flux distribution predicted for CAT production (Fig.~\ref{fig:flux}).
To investigate the influence of amino acid supplementation, we compared the simulated metabolic flux distributions for CAT production with and without external amino acids.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./figs/Fig-5-FluxDistribition-Optimal.pdf}
\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation, for three different cases: (A) amino acid uptake and synthesis, (B) amino acid uptake without synthesis, and (C) amino acid synthesis without uptake. Mean flux across the ensemble, normalized to glucose uptake flux. Thick arrows indicate flux to or from amino acids.}
\label{fig:flux}
\end{figure}
In the presence of amino acid supplementation, and \textit{de~novo} amino acid synthesis, there was an incomplete TCA cycle, where a combination
of glucose and amino acids powered protein expression (Fig.~\ref{fig:flux}A).
Glucose was consumed to produce acetyl-coenzyme A, and associated by-products,
while glutamate was converted to alpha-ketoglutarate which traveled to oxaloacetic acid and pyruvate for additional amino acid biosynthesis.
On the other hand, in the presence of amino acid supplementation, but without \textit{de~novo} amino acid biosynthesis, there was no TCA cycle flux.
In this case, ATP was produced by a combination of substrate level and oxidative phosphorylation, where
ubiquinone was regenerated via \textit{nuo} activity, without relying on succinate dehydrogenase in the TCA cycle (Fig.~\ref{fig:flux}B).
These first two cases where amino acids were available had similar performance, and their respective metabolic flux distributions had a 99\% correlation for all proteins. However, in the absence of amino acid supplementation (where all amino acids were synthesized \textit{de~novo} from glucose),
the energy efficiency and carbon yield decreased; in this case the TCA cycle was largely complete and there was diversion of metabolic flux into the Entner-Doudoroff pathway to produce NADPH (Fig.~\ref{fig:flux}C). [Mike: we need a link sentence to the experimentally constrained case.]

% to both cases where amino acids were avaialbe.
% This drop in performance is due to the burden of synthesizing amino acids, which require NADPH.
% This leads to the relatively high flux in the first step of the pentose phosphate pathway to generate NADPH, thus, less NADH is available for oxidative phosphorylation.
% The performance metrics and sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
% %Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.

@@

\subsubsection{Potential Alternative Metabolic Optima}
Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm})
For example, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.
A knockout of the Entner-Doudoroff pathway has no effect on the norm productivity of CAT compared to no knockouts (Fig.~\ref{fig:norm}A).
In addition, pairwise knockouts of Entner-Doudoroff and other subgroups (i.e. pentose phosphate pathway, cofactors, folate metabolism, etc.) in the network result in the same optimal solution of CAT productivity.
However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B); ssFBA will reroute the flux to optimize the objective function.
Interestingly, there are crucial subgroups that are detrimental to CAT productivity.
For example, a single group knockout of glycolysis/gluconoeogenesis or oxidative phosphorylation leads to no CAT production.
There were also cases of group knockouts that had no effect on productivity or its flux distribution such as isoleucine, leucine, and valine biosynthesis knocked out with histidine biosynthesis.
Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
In our case, ssFBA reached the same optimal solution of CAT productivity for 39.9\% of the pairwise group knockouts, but had a different flux distribution than compared to no knockouts.
To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.
It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
\begin{figure}[t!]
\includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribition-Analysis.pdf}
\caption{The norm effect of pairwise knockouts of subgroups in the cell-free network (A) CAT productivity norm compared to no knockouts in the experimentally constrained case. (B) Flux distribution norm compared to no knockouts in the experimentally constrained case.}
\label{fig:norm}
\end{figure}

\subsection{Summary and conclusions}
In this study, we developed a sequence specific constraint based modeling approach to predict the performance of cell free protein synthesis reactions for a range of proteins. We showed first principle predictions for protein production of deGFP and CAT were in agreement with experimental measurements, under two different promoters and two different cell-free extract systems.
This modeling approach suggested correlations for productivity, energy efficiency and carbon yield as a function of the size of the protein.
Furthermore, global sensitivity analysis identified oxygen uptake as being instrumental for maintaining a high energy efficiency and carbon yield.
The translation rate was identified as the rate limiting step for productivity.
The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize energy efficiency and carbon yield.
In conclusion, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.

@@

\subsection*{Formulation and solution of the model equations.}
The sequence-specific flux balance analysis problem was formulated as a linear program:
\begin{equation}
 \begin{multlined}
	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{X} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
	\mathrm{Subject \; to:}
	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
\mathcal{L}_{i} \leq w_i \leq \mathcal{U}_{i}  \qquad i=1,2,\hdots,\mathcal{R}
 \end{multlined}
\end{equation}
where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective cost vector
and $\mathcal{L}_{i}$ and $\mathcal{U}_{i}$ denote the lower and upper bounds on flux $w_{i}$, respectively.
The transcription (T) and translation (X) stoichiometry was modeled based upon the template reactions of Allen and Palsson \cite{Allen:2003aa}:
\begin{table}[!]
\centering
    \caption{Transcription and translation template reactions for protein production.}
    \renewcommand{\arraystretch}{1.2}
    \resizebox{\textwidth}{!}{\begin{tabular}{llcl} \toprule
        \textbf{Description} & \multicolumn{3}{l}{\textbf{Template reaction}} \\ \toprule
        Transcription initiation & $G_\mathcal{P}+R_{T}$ & $\longrightarrow$& $G_{\mathcal{P}}^{*}$ \\
        Transcription ($w_{T}$) & $G_{\mathcal{P}}^{*}+\quad\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right)$ & $\longrightarrow$& $mRNA+G_{\mathcal{P}}+R_{T}+\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}} \eta_{k} \cdot PPi$ \\
        mRNA degradation &  $mRNA$ &$\longrightarrow$& \quad $\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left\{k\right\}MP$ \\
	Translation initiation & $mRNA+R_{X}$ &$\longrightarrow$& $R_{X}^{*}$ \\
        tRNA charging & $\alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big)$ &$\longrightarrow$& $\alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)$ \\
 	&&& $\qquad \qquad \qquad \qquad \quad  {j=1,2,\hdots,20}$\\
        Translation ($w_{X}$) &  $R_{X}^{*}+\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big)$ &$\longrightarrow$& $\mathcal{P}+R_{X}+mRNA$\\
&&& $\quad +\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)$ \\ \bottomrule
    \end{tabular}}
\label{tbl:parameters}
\end{table}
where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
$R_{T}$ denotes the concentration of RNA polymerase,
$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase (open complex),
$\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
$Pi$ denotes inorganic phosphate,
$R_{X}$ denotes the ribosome concentration,
$R_{X}^{*}$ denotes bound ribosome,
and $AA_{j}$ denotes $j^{th}$ amino acid.
